How Slow do my Transient Typechecks Go?

### 2019-08-14 paper.tex
Make various typo fixes

Our work is based on Moth virtual machine 
\cite{Roberts2017,roberts-and-co-ecoop-2019}
an implementation
of the Grace programming language 
\citep{graceOnward12,graceSigcse13},
based on the Graal and Truffle toolchain
\cite{Wurthinger:2017:PPE,Wurthinger2013}
and developed from a Newspeak implementation based on the  Simple
Object Machine \cite{Daloze2016,SOMns}.

Grace is an object-oriented, imperative, educational programming
language, with a focus on introductory programming
courses, but also intended for more advanced study and research \citep{graceOnward12,graceSigcse13}.
%
While Grace's syntax draws
from the so-called ``curly bracket'' tradition of C, Java, and
JavaScript, the structure of the language
is in many ways closer to Smalltalk:
all computation is via dynamically dispatched  ``method requests''
where the object receiving the request decides which code to run,
and
%
returns within lambdas that are ``non-local'', returning to the method
activation in which the block is instantiated \citep{bluebook}.  In
other ways, Grace is closer to JavaScript than Smalltalk: Grace
objects can be created from object literals, rather than by
instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
objects and classes can be deeply nested within each 
other \citep{betabook}.

@@

The static core of Grace's type system is well described
elsewhere\citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object implements a type whenever it
implements all the methods required by that type,
rather than requiring classes or objects to declare types explicitly.
Grace's transient dynamic typechecks are first order, so methods match when they have
the same name and arity: 
argument and return types are ignored.
A type thus expresses the requests an object can respond to,
for example whether a particular accessor is available,
rather than a nominal location in an explicit inheritance hierarchy.

Grace then checks the types of values at run time:
%
\begin{itemize}
\item the values of arguments are checked after a method is requested, 
      but before the body of the method is executed;
\item the value returned by a method is checked after its body is executed; and
\item the values of variables are checked
      whenever written or read by user code.
\end{itemize}

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth run programs with
transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mix typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark. where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program.,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
up to 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

@@

This is preliminary work: there are a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness run on the same implementation is
subject to the same issues. Regarding external validity, 
Moth is built
on the Truffle and Graal toolchain, so we expect
to other Graal
VMs doing similar AST-based optimizations of transient typechecks.
Because we rely on common techniques, 
we expect our results should be transferable to JIT implementations
more widely.

>>>>

Our work is based on the Moth virtual machine 
\cite{Roberts2017,roberts-and-co-ecoop-2019},
an implementation
of the Grace programming language 
\citep{graceOnward12,graceSigcse13},
based on the Graal and Truffle toolchain
\cite{Wurthinger:2017:PPE,Wurthinger2013}
and developed from a Newspeak implementation based on the  Simple
Object Machine \cite{Daloze2016,SOMns}.

Grace is an object-oriented, imperative, educational programming
language, with a focus on introductory programming
courses, but also intended for more advanced study and research \citep{graceOnward12,graceSigcse13}.
%
While Grace's syntax draws
from the so-called ``curly bracket'' tradition of C, Java, and
JavaScript, the structure of the language
is in many ways closer to Smalltalk:
all computation is via dynamically dispatched  ``method requests''
where the object receiving the request decides which code to run,
and
%
returns within lambdas are ``non-local'', returning to the method
activation in which the block is instantiated \citep{bluebook}.  In
other ways, Grace is closer to JavaScript than Smalltalk: Grace
objects can be created from object literals, rather than by
instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
objects and classes can be deeply nested within each 
other \citep{betabook}.

@@

The static core of Grace's type system is well-described
elsewhere\citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object implements a type whenever it
implements all the methods required by that type,
rather than requiring classes or objects to declare types explicitly.
% This should say Moth's, not Grace's, but I know there is objection to that
Grace's transient dynamic typechecks are first-order, so methods match when they have
the same name and arity: 
argument and return types are ignored.
A type thus expresses the requests an object can respond to,
for example whether a particular accessor is available,
rather than a nominal location in an explicit inheritance hierarchy.

% Ditto
Grace then checks the types of values at run time:
%
\begin{itemize}
\item the values of arguments are checked after a method is requested, 
      but before the body of the method is executed;
\item the value returned by a method is checked after its body is executed; and
\item the values of variables are checked
      whenever written or read by user code.
\end{itemize}

@@

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth runs programs with
transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mix typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark, where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per-module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program.,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
these kinds of languages by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
up to 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

@@

This is preliminary work: there are a number of threats to validity. 
Regarding construct validity, 
our underlying
implementation may contain undetected bugs that affect the semantics
or performance of the gradual typing checks. Regarding internal
validity,
our benchmarking harness run on the same implementation is
subject to the same issues. Regarding external validity, 
Moth is built
on the Truffle and Graal toolchain, so we expect
to resemble other Graal
VMs doing similar AST-based optimizations of transient typechecks.
Because we rely on common techniques, 
we expect our results should be transferable to JIT implementations
more widely.
